Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant

### 2021-02-10 analysis/paper/paper.Rmd
finished draft of figures and added to written results.

institute:
  - UF: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
linkcolor: blue
bibliography: references.bib
csl: "../templates/ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
highlights: |
  These are the highlights
---

@@

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

@@

$$
g[\textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-24}) \\
\alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \textrm{for plot }j
$$

@@

@@

Overall survival rate across the entire dataset is 0.95 in both CF and 1-ha.
Survival was lowest in 2004 which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004.
Survival was lower in continuous forest compared to 1-ha fragments in this year (\~0.91 compared to \~0.94, will get exact numbers from data).
The lowest survival for 1-ha fragment plots (about 0.93) was actually in the 2006 census corresponding to unusually wet weather (SPEI \>0.5), but with a moderate drought the preceding October.

Log transformed plant size had a significant effect on survival in both continuous forest and 1-ha fragments (both p \< 0.001).
Survival probability of large plants approached 1, but this high survival rate was reached at a smaller size for plants in continuous forest compared to fragments (Figure \_\_\_ <!--# might be good to plot both curves on same graph -->).
Survival of the smallest plants was higher in 1-ha fragments, but the confidence intervals still overlapped the fitted values for continuous forest.

Drought history had a significant effect on survival in both habitats (P \< 0.001), but the shape of the crossbasis surface differed.
For 1-ha fragments, recent SPEI values had the strongest effect on survival with the highest survival near SPEI of 0 and decreasing survival in both drier and wetter conditions (Figure \_\_\_).
In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 producing slightly reduced survival probabilities.
However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year, also effected survival probabilities.
Lower SPEI (drought) 15--20 months prior to census was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
Plants in both continuous forest and 1-ha fragments showed an increase in survival probability with very high (wet) SPEI values at a lag time of 36 months.
However, only the first year of census data (1999) had an SPEI value greater than 2 within 36 months in the past, and although there were plants that did not survive this initial census, it may not be representative of a true survival process <!--# Emilio: let's talk about this. Maybe I should remove this year from the analysis for survival? -->

@@

Plants were on average larger in continuous forest compared to 1-ha fragments.
In continuous forest, plants had an average of 2.91 shoots and were on average 40.6 cm tall with an average size (shoots $\times$ height) of 150.
In 1-ha fragments, plants had 2.53 shoots on average and a mean height of 36.2 cm and a mean size of 115.
Log transformed size in year t was a significant predictor of log transformed size in year t+1 (p \< 0.001) and the relationship was close to linear for both continuous forest (edf = ) and 1-ha fragments (edf = ).

The size difference between plants in continuous forest and 1-ha fragments was more dramatic early on in the dataset with mean sizes becoming more similar over time.

@@

### Continuous Forest

size: p \< 0.001

plot: p \< 0.001

SPEI: p \< 0.001

Predicted results show only lagged effects of SPEI on growth (f(x) at lags 0 and 1 is flat).
Wet conditions in preceding winter months improve growth (lags 4-2, Oct - Dec).
Positive effect of drought on growth at lags 11-8 (Mar - Jun) and 34-24 (nearly the entirety of year t-2).

### 1-ha fragment

size: p \< 0.001

plot: p \< 0.001

SPEI: p \< 0.001

Crossbasis looks very similar to continuous forest but more extreme.
SPEI has a stronger effect on growth in fragments.
Wet conditions at lag 0 predict the greatest shrinkage of plants in fragments (∆size = -75 at SPEI \>2), but at lags 5-2 (Sept-Dec) high SPEI has a less negative effect on plant size.
We only expect plant growth (∆size \> 0) in years when SPEI was low at lags 34-32

The size of plants significantly predicted their size in the following year in continuous forest (edf = 7.8, P \< 0.001 <!--# What stats to include?  reference df and F also?  Is this all better as a table? -->) and forest fragments (edf = 7.4, P \< 0.001).
In continuous forest, plants under a size of 259 were predicted to grow from year t-1 to year t, and the fit line was below 1:1 for larger plants, indicating that larger plants tended to shrink, on average.
In 1 ha fragments, the fit line followed a 1:1 line more closely, indicating that plants tended to stay the same size on average from year to year over the course of this study.
<!--# something weird about backtransformation causing these strange results? --> Drought history also had a significant effect on plant size in both continuous forest (edf = 14.7, p \< 0.001) and fragments (edf = 17.8, p \< 0.001).

@@

Probability of flowering increased with size and was generally higher in continuous forest compared to 1-ha fragments.
The exceptions to this are in 2006 and 2008 when the proportion of flowering plants was lower in continuous forest.
The most striking difference between habitat types in flowering proportion was in 2003 when 9.7% of plants flowered in the continuous forest plots and only 2.7% flowered in fragment plots.

### Continuous forest

size: p \< 0.001

plot: p\<0.001

SPEI: p \< 0.001

Flowering probability was increased by recent drought (at lag = 0 with SPEI \< 1) and drought at lags 20-15 (Jun - Nov).
Drought at lags 13-7 reduced flowering probability

## Size by year interaction

<!--# this all needs updating.  Probably going to use AIC instead of these p-values since I think they're doing the same thing. -->

In continuous forest plots there was support for size by year interactions for growth (95% LR~null~ = 4.54, LR~obs~ = 45.69) and survival (95% LR~null~ = 3.91, LR~obs~ = 29.18) , but not for flowering (95% LR~null~ = 4.44, LR~obs~ = 1.77).
However, in 1 ha fragments, we saw no evidence of size by year interaction.
For growth 95% LR~null~ = 4.45, LR~obs~ = 4.35, for survival 95% LR~null~ = 4.47, LR~obs~ = 0.62, and for flowering probability 95% LR~null~ = 3.86, LR~obs~ = 2.82.
This is consistent with the DLNMs in that the crossbasis functions for \_\_\_ and \_\_\_ were much flatter in fragments compared to continuous forest.

>>>>

institute:
  - UF: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: no
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
linkcolor: blue
bibliography: references.bib
csl: "../templates/ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
highlights: |
  These are the highlights
---

@@

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
library(knitr)
library(here)
```

@@

$$
\begin{gather}
    g[ \textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-24}) \\
    \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \textrm{for plot }j 
\end{gather}
$$

@@

```{r}
cap <- "Plots of means ± SE."
```

```{r eda, fig.cap=cap}
knitr::include_graphics(here("analysis", "figures", "eda.png"))

```

@@

```{r}
cap <- "Smooth effect of plant size in the previous census on survival modeled by a cubic regression spline.  95% confidence intervals are shown and include uncertainty in the model intercept as well as uncertainty due to smoothness selection. The smooths for 1-ha fragments and continuous forest are fit in separate models. The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale"
```

```{r surv-size, fig.cap=cap}
include_graphics(here("analysis", "figures", "surv_both_size.png"))
```

Overall survival rate across the entire dataset is 0.95 in both CF and 1-ha.
Survival was lowest in 2004 which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004 (Figure \@ref(fig:eda)).
Survival was lower in continuous forest compared to 1-ha fragments in this year (\~0.91 compared to \~0.94, <!--# will get exact numbers from data -->).
The lowest survival for 1-ha fragment plots (about 0.93) was actually in the 2006 census corresponding to unusually wet weather (SPEI \>0.5), but with a moderate drought the preceding October.

Log transformed plant size had a significant effect on survival in both continuous forest and 1-ha fragments (both p \< 0.001).
Survival probability of large plants approached 1, but this high survival rate was reached at a smaller size for plants in continuous forest compared to fragments (Figure \@ref(fig:surv-size)).
Survival of the smallest plants was higher in 1-ha fragments, but the 95% confidence intervals still overlapped the fitted values for continuous forest <!--# I wonder if this changes if I don't include both sources of uncertainty in draw.gam? -->.

```{r}
cap <- "Smooth effect of lagged SPEI on survival. Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis. Contour lines correspond to a change of 0.002. Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale."

```

```{r surv-spei, fig.cap=cap}
knitr::include_graphics(here("analysis", "figures", "surv_both_spei.png"))
```

Drought history had a significant effect on survival in both habitats (P \< 0.001), but the shape of the crossbasis surface differed.
For 1-ha fragments, recent SPEI values had the strongest effect on survival with the highest survival near SPEI of 0 and decreasing survival in both drier and wetter conditions (Figure \@ref(fig:surv-spei)a).
In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 producing slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year, also effected survival probabilities.
Lower SPEI (drought) 15--20 months prior to census was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
Plants in both continuous forest and 1-ha fragments showed an increase in survival probability with very high (wet) SPEI values at a lag time of 36 months.
However, only the first year of census data (1999) had an SPEI value greater than 2 within 36 months in the past, and although there were plants that did not survive this initial census, it may not be representative of a true survival process <!--# Emilio: let's talk about this. Maybe I should remove this year from the analysis for survival? -->

@@

Plants were on average larger in continuous forest compared to 1-ha fragments.
In continuous forest, plants had an average of 2.91 shoots and were on average 40.6 cm tall with an average size (shoots $\times$ height) of 150.
In 1-ha fragments, plants had 2.53 shoots on average and a mean height of 36.2 cm and a mean size of 115.
The size difference between plants in continuous forest and 1-ha fragments was more dramatic early on in the timeseries with mean sizes becoming more similar over time (Figure \@ref(fig:eda)a).

```{r}
cap <- "Smooth effect of plant size in year t on plant size in year t+1 modeled by a cubic regression spline.  95% confidence intervals are shown and include uncertainty in the model intercept as well as uncertainty due to smoothness selection. The dashed red line represents a 1:1 line for reference. The smooths for 1-ha fragments and continuous forest are fit in separate models. Model intercepts were added to the fitted values of the smooths."
```

```{r growth-size, fig.cap=cap}
include_graphics(here("analysis", "figures", "growth_both_size.png"))
```

Log transformed size in year t was a significant predictor of log transformed size in year t+1 (p \< 0.001).
The 95% confidence intervals for the effect of size in year t in fragments and continuous forest overlapped for most sizes, but was slightly higher in continuous forest for mid-sized plants (Figure \@ref(fig:growth-size)).

```{r}
cap <- "Smooth effect of lagged SPEI on plant growth. Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis. Contour lines correspond to a change of 0.1. "
```

<!--# this is the raw fitted values of the tensor product smooth. That is, I have not shifted this by adding the model intercepts like I did in the last figure.  Not sure if the effect size should actually be interpreted as a additive change in size, but I think so maybe?  If that's true, then the predicted effect of SPEI on size in year t+1 is extremely small compared to the effect of size in year t. -->

```{r growth-spei, fig.cap=cap}
include_graphics(here("analysis", "figures", "growth_both_spei.png"))
```

Drought history also had a significant effect on plant size in both habitats (p \< 0.001).
The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 5.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
Drought at lags 8--10 increase growth in both habitats.
<!--# say a lot more about this here. -->

The random effect of plot is significant in both habitat models (p \< 0.001).

@@

@@

Probability of flowering was generally higher in continuous forest compared to 1-ha fragments.
The exceptions to this are in 2006 and 2008 when the proportion of flowering plants was lower in continuous forest (Figure \@ref(fig:eda)c).
The most striking difference between habitat types in flowering proportion was in 2003 when 9.7% of plants flowered in the continuous forest plots and only 2.7% flowered in fragment plots.

```{r}
cap <- "Smooth effect of plant size in the previous census on flowering probability modeled by a cubic regression spline.  95% confidence intervals are shown and include uncertainty in the model intercept as well as uncertainty due to smoothness selection. The smooths for 1-ha fragments and continuous forest are fit in separate models. The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale"
```

```{r flwr-size, fig.cap=cap}
include_graphics(here("analysis", "figures", "flwr_both_size.png"))
```

Plant size had a significant effect on flowering probability in both habitats (p \< 0.001).
Flowering probability increased rapidly with log-transformed plant size greater than 5 in both habitats (Figure \@ref(fig:flwr-size)).
Flowering probability was greater in 1-ha fragments for large plants compared to continuous forest.
However, because mean plant size is larger in continuous forest, overall flowering probability was greater.
<!--# idea: plot vertical lines representing mean or median size in each habitat -->

```{r}
cap <- "Smooth effect of lagged SPEI on flowering probability. Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis. Contour lines correspond to a change of 0.001. Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale."
```

```{r flwr-spei, fig.cap=cap}
include_graphics(here("analysis", "figures", "flwr_both_spei.png"))
```

Drought history had a significant effect on flowering probability in both habitat types (p \< 0.001).
Flowering probability was overall higher at all values of SPEI in continuous forest compared to 1-ha fragments (Figure \@ref(fig:flwr-spei)).
The effects of SPEI on flowering probability were also stronger and more complex in continuous forest compared to fragments (1-ha edf = 8.5, continuous forest edf = 10.5).
In continuous forest, flowering probability was increased by recent drought (at lag = 0 with SPEI \< -1) and drought at lags 15--20 (Jun - Nov).
Drought at lags 7--13 reduced flowering probability.
The shape of the crossbasis function was similar in 1-ha fragments, but significantly muted (Figure \@ref(fig:flwr-spei)a).
That is, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.

The random effect of plot also had a significant effect in both habitats (p \< 0.001).
